---
title: "Test_Framework"
output: html_document
date: "2025-11-29"
---

# PURPOSE
# This notebook implements the TRAINING PIPELINE only (no external test set).
# It covers: data loading → preprocessing → individual filtering → forward selection → MINT modelling.
# Use Test_Framework(Reproducibility).Rmd to extend with the HGMA test set and external validation.

```{r}
# Clear the workspace to avoid stale variables from previous runs
rm(list=ls())
# Load all helper functions (preprocessing, filtering, modelling, visualisation)
source("Function_Helpers.R")
```

# Packages installation
```{r}
# req_pkgs: character vector of all packages required by this pipeline
# - dplyr:                   data manipulation (filter, select, mutate, bind_rows)
# - mixOmics:                PLS-DA, MINT sPLS-DA, CLR transformation, perf(), tune()
# - curatedMetagenomicData:  access to curated gut microbiome study datasets (CMD v3.16.1)
# - SummarizedExperiment:    data container used by curatedMetagenomicData (assay(), colData())
# - stringr:                 string manipulation (str_detect, str_sub, regex)
# - testthat:                unit testing utilities (used in validation checks)
# - tibble:                  tibble data frames (used by some helper functions)
# - purrr:                   functional programming (map, imap, map_lgl)
req_pkgs = c('dplyr', 'mixOmics', 'curatedMetagenomicData', 'SummarizedExperiment',
             'stringr', 'testthat', 'tibble', 'stringr', 'purrr',
             'openxlsx')  # openxlsx: required by run_mint() to export selected taxa + error rates to Excel

install_if_missing(req_pkgs)
```

## Data Loading
```{r}
# study_names: CMD query strings for the 5 training cohorts
# Format: "<AuthorYear>.<data_type>"
# The 5 studies span diverse populations (Chinese, Swedish, Indian, American):
#   FengQ_2015       – colorectal cancer cohort (Chinese); contains T2D/Healthy
#   KarlssonFH_2013  – Swedish T2D cohort; the canonical T2D microbiome study
#   LiJ_2014         – large-scale Chinese metagenome reference cohort
#   YuJ_2015         – colorectal cancer cohort (Chinese); also contains T2D/Healthy
#   SankaranarayananK_2015 – Indian cohort

# NOTE: HMP2019 (6th training study from the thesis) is NOT listed here yet;
# it may be added during full_framework.Rmd finalisation.
study_names <- c("FengQ_2015.relative_abundance", "KarlssonFH_2013.relative_abundance", "LiJ_2014.relative_abundance",
                 "YuJ_2015.relative_abundance", "SankaranarayananK_2015.relative_abundance")

# Load and process each study
# abundance_lst: named list (length 5); each element is a data.frame:
#   rows = samples (sample IDs as row names)
#   cols = taxa (species names as column names, e.g. "Bacteroides uniformis")
# counts=TRUE retrieves raw counts (not pre-normalised relative abundances)
# rownames="short" assigns short sample IDs rather than full accession strings
# t() transposes from taxa×samples to samples×taxa orientation
# check.names=FALSE preserves special characters and spaces in taxa names
abundance_lst <- lapply(study_names, function(study) {
  rawdata <- curatedMetagenomicData(study, dryrun = FALSE, counts = TRUE, rownames = "short")
  raw_tse <- rawdata[[1]]  # extract the SummarizedExperiment object
  abund <- assay(raw_tse, "relative_abundance")  # taxa × samples matrix
  
  # Joyce: added checknames =false to prevent issues with inconsistent colnames later
  as.data.frame(t(abund), check.names = FALSE)  # transpose → samples × taxa
})

## output data: keys are study name, each col is taxa name

# meta_lst: named list (length 5); each element is a data.frame:
#   rows = samples (sample IDs as row names)
#   cols = all CMD metadata columns (includes 'disease', 'gender', 'body_site', etc.)
# Extract metadata for each study
meta_lst <- lapply(study_names, function(study) {
  rawdata <- curatedMetagenomicData(study, dryrun = FALSE, counts = TRUE, rownames = "short")
  raw_tse <- rawdata[[1]]
  as.data.frame(colData(raw_tse), check.names = FALSE)  # sample metadata from SummarizedExperiment
})

```



```{r}
# study_names_v2: shortened study names (strip ".relative_abundance" suffix)
# e.g. "FengQ_2015.relative_abundance" → "FengQ_2015"
# Used as list names for abundance_lst and meta_lst so they are human-readable
# and match the format used as the study label in modelling.
study_names_v2 = c()
for (i in 1:length(strsplit(study_names, "\\.")) ) {
  # strsplit splits on "."; take the first element (the study name before the dot)
  study_names_v2[i] = strsplit(study_names, "\\.")[[i]][1]
}

# Apply shortened names to both lists (enables name-based access downstream)
names(abundance_lst) <- study_names_v2
names(meta_lst) <- study_names_v2
```


## Data Processing
```{r}
# keep_cols: additional metadata columns to retain alongside 'disease'
#   "gender"    – sex of the participant (used in downstream characterisation)
#   "body_site" – sampling location (all should be stool for gut microbiome studies)
keep_cols <-c("gender", "body_site")

# diseases: the two disease classes we are discriminating
#   "Healthy" – Normal Glucose Tolerance (NGT) / control participants
#   "T2D"     – Type 2 Diabetes participants
diseases <- c("Healthy", "T2D")

# run_preprocessing() is the main preprocessing entry point (Function_Helpers.R).
# For 5 studies it routes to preprocess_multiple() → preprocess_single() per study → integrate_filter()
#
# Parameters used:
#   percent = 0.01    → remove taxa contributing < 0.01% of total abundance (low-count removal)
#   offset_value = 1  → add 1 to all counts before CLR (standard for count data to handle zeros)
#   threshold = 1     → keep taxa present in at least 1 study (effectively: keep all)
#
# res: named nested list; keys = study short names (e.g. "FengQ_2015")
#   res[[study_name]]$abund – CLR-transformed abundance data frame (samples × surviving taxa)
#   res[[study_name]]$meta  – filtered metadata (disease, gender, body_site columns)
res<- run_preprocessing(abundance_list = abundance_lst, meta_list = meta_lst, diseases = diseases, keep_cols = keep_cols, percent = 0.01, offset_value = 1, threshold = 1)

```

# Model Development and Training
## Data Integration
### Individual filtering
```{r}
# individual_filtering() (Layer 2): removes studies where any disease class is too
# hard to discriminate (per-class BER > err_thres at ncomp components).
#
# INPUT:  res – preprocessed study list from run_preprocessing()
# OUTPUT: result_individual – filtered study list (same structure as res, failing studies removed)
#
# Parameters:
#   dist      = "centroids.dist" – nearest-centroid classification rule for PLS-DA CV
#   ncomp     = 2                – evaluate at 2 PLS-DA components
#   err_thres = 0.5              – discard any study where a class BER > 50%
#                                  (NOTE: thesis specifies 0.7; reconcile before full_framework.Rmd)
#
# CMD_filtered = readRDS("res")  # [ORIGINAL] was loaded from disk; now kept in memory
result_individual <- individual_filtering(res, dist="centroids.dist", ncomp=2, err_thres = 0.5)
print("Individual Filtering — selected studies:")
names(result_individual)  # prints the names of studies that passed the filter
```

### Forward selection
```{r}
# forward_selection() (Layer 3): greedy algorithm to find the optimal SUBSET of
# studies to integrate. Studies are added one-at-a-time; a study is accepted only
# if adding it reduces the multi-study MINT BER.
#
# INPUT:  res – preprocessed study list (NOTE: uses res, not the individually-filtered result;
#               forward selection implicitly handles noisy studies via BER)
# OUTPUT: result_forward – named list of the selected study objects (abund + meta)
#
# Parameters:
#   dist  = "centroids.dist" – classification rule for MINT evaluation
#   ncomp = 2                – number of MINT components
result_forward = forward_selection(res, dist="centroids.dist", ncomp=2)
print("Forward Selection — selected studies:")
names(result_forward)  # prints which studies were selected in the final combination

```

## Modelling
### Individual Filtering — MINT Model
```{r}
## Assemble integrative dataset from individually-filtered studies

# Add 'study' column to each study's metadata for provenance tracking after bind_rows
for (name in names(result_individual)) {
  result_individual[[name]]$meta$study = strsplit(name, '\\.')[[1]][1]
}

# result_df_individual: all individually-filtered studies stacked into one data frame
# $abund – combined CLR abundance (rows = all samples, cols = taxa)
# $meta  – combined metadata (rows = all samples, includes $study column)
result_df_individual = bind_rows(result_individual)

# p_selected_individual: training abundance matrix for the individual-filtering strategy.
# Used in align_columns() when aligning with the HGMA test set in Reproducibility notebook.
p_selected_individual <- result_df_individual$abund

# Prepare run_mint() inputs for the individual-filtering strategy
# study must exist in the calling environment — run_mint() reads it directly
X     <- result_df_individual$abund
Y     <- as.factor(result_df_individual$meta$disease)
study <- as.factor(result_df_individual$meta$study)

# run_mint() with label="individual" → outputs saved as:
#   output_files/individual_selected_taxa.xlsx
#   (plots saved by plot_mint_results below)
mint_results_individual <- run_mint(X = X, Y = Y, ncomp = 2, dist = "centroids.dist",
                                    label = "individual")

# plot_mint_results() with label="individual" → saves:
#   output_files/individual_scoreplot.png
#   output_files/individual_loadings_comp1.png
#   output_files/individual_loadings_comp2.png
plot_mint_results(mint_results_individual, Y = Y, study = study, label = "individual")
```

```{r}
## Confusion matrix — Individual Filtering training set
# predict() on training data gives in-sample performance (optimistic upper bound).
predict_individual <- predict(mint_results_individual$model, newdata = X,
                              dist = "centroids.dist", study.test = study)

# conf_mat_individual_c1/c2: confusion matrices at component 1 and component 2
conf_mat_individual_c1 = get.confusion_matrix(truth = result_df_individual$meta$disease,
                                              predicted = predict_individual$class$centroids.dist[, 1])
conf_mat_individual_c2 = get.confusion_matrix(truth = result_df_individual$meta$disease,
                                              predicted = predict_individual$class$centroids.dist[, 2])
conf_mat_individual_c1
conf_mat_individual_c2
```

### Forward Selection — MINT Model
```{r}
## Assemble integrative dataset from forward-selected studies

# Add 'study' column to each study's metadata for provenance tracking after bind_rows
for (name in names(result_forward)) {
  result_forward[[name]]$meta$study = strsplit(name, '\\.')[[1]][1]
}

# result_df_forward: all forward-selected studies stacked into one data frame
# $abund – combined CLR abundance (rows = all samples, cols = taxa)
# $meta  – combined metadata (rows = all samples, includes $study column)
result_df_forward = bind_rows(result_forward)

# p_selected_forward: training abundance matrix for the forward-selection strategy.
# Used in align_columns() when aligning with the HGMA test set in Reproducibility notebook.
p_selected_forward <- result_df_forward$abund

# Prepare run_mint() inputs for the forward-selection strategy
# study must exist in the calling environment — run_mint() reads it directly
X     <- result_df_forward$abund
Y     <- as.factor(result_df_forward$meta$disease)
study <- as.factor(result_df_forward$meta$study)

# run_mint() with label="forward" → outputs saved as:
#   output_files/forward_selected_taxa.xlsx
#   (plots saved by plot_mint_results below)
mint_results_forward <- run_mint(X = X, Y = Y, ncomp = 2, dist = "centroids.dist",
                                 label = "forward")

# plot_mint_results() with label="forward" → saves:
#   output_files/forward_scoreplot.png
#   output_files/forward_loadings_comp1.png
#   output_files/forward_loadings_comp2.png
plot_mint_results(mint_results_forward, Y = Y, study = study, label = "forward")
```

```{r}
## Confusion matrix — Forward Selection training set
# predict() on training data gives in-sample performance (optimistic upper bound).
predict_forward <- predict(mint_results_forward$model, newdata = X,
                           dist = "centroids.dist", study.test = study)

# conf_mat_forward_c1/c2: confusion matrices at component 1 and component 2
conf_mat_forward_c1 = get.confusion_matrix(truth = result_df_forward$meta$disease,
                                           predicted = predict_forward$class$centroids.dist[, 1])
conf_mat_forward_c2 = get.confusion_matrix(truth = result_df_forward$meta$disease,
                                           predicted = predict_forward$class$centroids.dist[, 2])
conf_mat_forward_c1
conf_mat_forward_c2
```




```{r}
# [ORIGINAL] Diagnostic dim checks — redundant now that X, Y, study are assigned above before run_mint()
# X = result_df$abund; Y = as.factor(result_df$meta$disease)
# dim(result_df$abund)
# dim(as.data.frame(result_df$meta$disease))
# dim(as.data.