---
title: "Test_Framework"
output: html_document
date: "2025-11-29"
---

```{r}
rm(list=ls()) 
source("Function_Helpers.R")
```

# Packages installation
```{r}
req_pkgs = c('dplyr', 'mixOmics', 'curatedMetagenomicData', 'SummarizedExperiment', 
             'stringr', 'testthat', 'tibble', 'stringr', 'purrr')

install_if_missing(req_pkgs) 
```

## Data Loading
```{r}
study_names <- c("FengQ_2015.relative_abundance", "HMP_2019_t2d.relative_abundance", 
                 "KarlssonFH_2013.relative_abundance", "LiJ_2014.relative_abundance", 
                 "YuJ_2015.relative_abundance", "SankaranarayananK_2015.relative_abundance")

# Load and process each study
abundance_lst <- lapply(study_names, function(study) {
  rawdata <- curatedMetagenomicData(study, dryrun = FALSE, counts = TRUE, rownames = "short")
  raw_tse <- rawdata[[1]]
  abund <- assay(raw_tse, "relative_abundance")
  # Joyce: added checknames =false to prevent issues with inconsistent colnames later
  as.data.frame(t(abund), check.names = FALSE)
})

## output data: keys are study name, each col is taxa name

# Extract metadata for each study
meta_lst <- lapply(study_names, function(study) {
  rawdata <- curatedMetagenomicData(study, dryrun = FALSE, counts = TRUE, rownames = "short")
  raw_tse <- rawdata[[1]]
  as.data.frame(colData(raw_tse), check.names = FALSE)
})
```

```{r}
# feature alignment for taxa names
align_columns <- function(p_test, p_selected) {
  
  # Remove 'species:' prefix from column names
  colnames(p_selected) <- sub("^species:\\s*", "", colnames(p_selected), ignore.case = TRUE)
  colnames(p_test) <- sub("^species:\\s*", "", colnames(p_test), ignore.case = TRUE)
  
  change_log <- list()
  message_log <- character()  # Collect messages for single matches
  
  for (i in seq_along(colnames(p_test))) {
    
    col_test <- colnames(p_test)[i]
    
    # Only process test columns with '/'
    if (grepl("/", col_test, fixed = TRUE)) {
      
      taxa_test <- trimws(unlist(strsplit(col_test, "/", fixed = TRUE)))
      matched_train_cols <- c()
      
      # Find matching train columns
      for (col_train in colnames(p_selected)) {
        taxa_train <- trimws(unlist(strsplit(col_train, "/", fixed = TRUE)))
        if (all(taxa_train %in% taxa_test)) {
          matched_train_cols <- c(matched_train_cols, col_train)
        }
      }
      
      # Handle matches
      if (length(matched_train_cols) == 1) {
        old_name <- col_test
        colnames(p_test)[i] <- matched_train_cols[1]
        change_log[[old_name]] <- paste0("Single match: renamed test column to '", matched_train_cols[1], "'")
        
        # Add to message log
        message_log <- c(
          message_log,
          paste0("Single match found: test column '", old_name, "' renamed to '", matched_train_cols[1], "'")
        )
        
      } else if (length(matched_train_cols) > 1) {
        new_train_col <- paste(matched_train_cols, collapse = "/")
        
        # Create combined column in p_selected if needed
        if (!(new_train_col %in% colnames(p_selected))) {
          p_selected[[new_train_col]] <- rowSums(p_selected[, matched_train_cols, drop = FALSE])
          change_log[[paste0("p_selected: ", new_train_col)]] <- paste0(
            "Created new combined column from: ", paste(matched_train_cols, collapse = ", ")
          )
        }
        
        old_name <- col_test
        colnames(p_test)[i] <- new_train_col
        change_log[[old_name]] <- paste0(
          "Multiple matches: renamed test column to '", new_train_col, "'"
        )
      }
      # No matches → leave unchanged
    }
  }
  
  # Align training and testing datasets
  common_cols <- intersect(colnames(p_selected), colnames(p_test))
  training <- p_selected[, common_cols, drop = FALSE]
  testing <- p_test[, common_cols, drop = FALSE]
  
  # Combine message log into a single variable
  all_messages <- paste(message_log, collapse = "\n")
  
  return(list(training = training, testing = testing, change_log = change_log, messages = all_messages))
}
```

```{r}
# change the study name so users can read it easily.
study_names_v2 = c()
for (i in 1:length(strsplit(study_names, "\\.")) ) {
  study_names_v2[i] = strsplit(study_names, "\\.")[[i]][1]
}

names(abundance_lst) <- study_names_v2  
names(meta_lst) <- study_names_v2
```


## Data Processing
```{r}
keep_cols <-c("gender", "body_site")
diseases <- c("Healthy", "T2D")
res<- run_preprocessing(abundance_list = abundance_lst, meta_list = meta_lst, diseases = diseases, keep_cols = keep_cols, percent = 0.01, offset_value = 1, threshold = 1)
```

# Model Development and Training
## Data Integration
### Individual filtering
```{r}
# CMD_filtered = readRDS("res")
result <- individual_filtering(res, dist="centroids.dist", ncomp=2, err_thres = 0.5)
names(result)
```

## getting combined rawdata based on taxa from CMD filtering and studies from forward selection 
```{r}
library(dplyr)

# extract study names
studies <- names(result)

# subset raw abundance data per study
subset_raw_lst <- lapply(studies, function(study) {

  # processed abundance (from previous function)
  abund_proc <- result[[study]]$abund

  # raw abundance data
  abund_raw <- abundance_lst[[study]]

  # rows (samples) and columns (taxa) to keep
  rows_keep <- rownames(abund_proc)
  cols_keep <- colnames(abund_proc)

  # strict checks: all must exist
  stopifnot(
    all(rows_keep %in% rownames(abund_raw)),
    all(cols_keep %in% colnames(abund_raw))
  )

  # subset raw data by rows AND columns
  abund_raw[
    rows_keep,
    cols_keep,
    drop = FALSE
  ]
})

# bind all studies by rows
p_selected <- bind_rows(subset_raw_lst)
```

## testing implementation of taxa renmaing 
```{r}
p_test <- readRDS("p_test")
align_columns(p_test, p_selected)
```

## raw code testing 
```{r}
p_test <- readRDS("p_test")
colnames(p_selected) <- sub("^species:\\s*", "", colnames(p_selected), ignore.case = TRUE)
# 1: Initialize message log
change_log <- list()

# 2: Loop over columns of p_test
for (i in seq_along(colnames(p_test))) {
  
  col_test <- colnames(p_test)[i]
  
  # Only process columns that contain '/'
  if (grepl("/", col_test, fixed = TRUE)) {
    
    # Step 2a: Split test column into taxa and trim whitespace
    taxa_test <- trimws(unlist(strsplit(col_test, "/", fixed = TRUE)))
    
    # Step 2b: Find matching train columns (p_selected)
    matched_train_cols <- c()
    
    for (col_train in colnames(p_selected)) {
      taxa_train <- trimws(unlist(strsplit(col_train, "/", fixed = TRUE)))  # works if no '/'
      
      if (all(taxa_train %in% taxa_test)) {
        matched_train_cols <- c(matched_train_cols, col_train)
      }
    }
    
    # Step 2c: Handle matches
    if (length(matched_train_cols) == 1) {
      # Single match → rename p_test column
      old_name <- col_test
      colnames(p_test)[i] <- matched_train_cols[1]
      change_log[[old_name]] <- paste0("Single match: renamed test column to '", matched_train_cols[1], "'")
      
      # Print only for single matches
      cat("\nSingle match found:\n")
      cat("  Test column:", old_name, "\n")
      cat("  Matched train column:", matched_train_cols[1], "\n")
      cat("  Action: renamed test column to", matched_train_cols[1], "\n")
      
    } else if (length(matched_train_cols) > 1) {
      # Multiple matches → create combined column in p_selected
      new_train_col <- paste(matched_train_cols, collapse = "/")
      
      # Only create new column if it does not already exist
      if (!(new_train_col %in% colnames(p_selected))) {
        p_selected[[new_train_col]] <- rowSums(p_selected[, matched_train_cols, drop = FALSE])
        change_log[[paste0("p_selected: ", new_train_col)]] <- paste0(
          "Created new combined column from: ", paste(matched_train_cols, collapse = ", ")
        )
      }
      
      # Rename p_test column to match new_train_col
      old_name <- col_test
      colnames(p_test)[i] <- new_train_col
      change_log[[old_name]] <- paste0(
        "Multiple matches: renamed test column to '", new_train_col, "'"
      )
    }
    
    # If no matches, leave column unchanged (no print)
  }
}

# 3: Outputs
cat("\n---\nFinal column names in p_test:\n")
print(colnames(p_test))

cat("\nFinal column names in p_selected:\n")
print(colnames(p_selected))

cat("\nChange log:\n")
print(change_log)
```













### Forward selection
```{r}
result = forward_selection(res, dist="centroids.dist", ncomp=2)
```

## Modelling
```{r}
## compose off integrative datasets for all selected studies
for (name in names(result)) {
  result[[name]]$meta$study = strsplit(name, '\\.')[[1]][1]
  # lst[[name]]$meta$study = name
}
# binding together
result_df = bind_rows(result)
# saveRDS here for schema level feature alignment
saveRDS(result_df$abund, "p_selected")

ab.mint.splsda = tune(method = "mint.splsda", X = result_df$abund, Y = as.factor(result_df$meta$disease), ncomp = 2, folds=5, nrepeat=10,
                      study = as.factor(result_df$meta$study), test.keepX = seq(10, ncol(result_df$abund), 1), validation = 'Mfold', dist = "centroids.dist") 
plot(ab.mint.splsda)
optimal.keepX = ab.mint.splsda$choice.keepX
optimal.keepX
mint.splsda.model = mint.splsda(X=result_df$abund, Y=result_df$meta$disease, 
                                study=result_df$meta$study, ncomp=2, keepX = optimal.keepX)

perf(mint.splsda.model, validation="Mfolds", folds=5, nrepeat=10, dist="centroids.dist")$global.error$error.rate.class

## Confusion matrix for training set
predict.splsda.null <- predict(mint.splsda.model, newdata = result_df$abund, 
                             dist = "centroids.dist", study.test = result_df$meta$study)

conf_mat_c1 = get.confusion_matrix(truth = result_df$meta$disease, predicted = predict.splsda.null$class$centroids.dist[, 1])
conf_mat_c2 = get.confusion_matrix(truth = result_df$meta$disease, predicted = predict.splsda.null$class$centroids.dist[, 2])
conf_mat_c1
conf_mat_c2
```


```{r}
X = result_df$abund; Y = as.factor(result_df$meta$disease)
# result_df$meta$study

dim(result_df$abund)
dim(as.data.frame(result_df$meta$disease))
dim(as.data.frame(result_df$meta$study))
```

### introducing testing set
```{r}
# testset1 = draw_TestSet(dataset_study="PRJEB1786")
# testset2 = draw_TestSet(dataset_study="PRJNA422434")
testset1_obj = readRDS("PRJEB1786.rds")
#testset2_obj = readRDS("PRJNA422434.rds")
keep_cols <-c("Gender", "sample.ID")
diseases <- c("Healthy", "T2D", "NGT")

# calculate the second smallest value
test_obj = testset1_obj
x <- unlist(test_obj$abund)
x <- x[x > 0]
second_smallest <- sort(unique(x), na.last = NA)[2]


#abundance_lst_test = list(test1 = test_obj$abund)
#meta_lst_test = list(test1 = test_obj$meta)

## instead of run_preprocessing we loop through pre_process single and then integrate filter to get counts without CLR
testset <- preprocess_single(meta=testset1_obj$meta, abund = testset1_obj$abund, diseases = diseases, keep_cols = keep_cols, percent = 0.01, offset_value = second_smallest, CLR=FALSE)



#testset<- run_preprocessing(abundance_list = abundance_lst_test, meta_list = meta_lst_test, diseases = diseases, keep_cols = keep_cols, percent = 0.01, offset_value = second_smallest, threshold = 1)




common_taxa = intersect( names(result_df$abund), names(testset$data) )
# add the renaming logic here with edits
# common_taxa = intersect( names(result_df$abund), names(testset$data) )
new_train = result_df$abund[, common_taxa, drop=F]
new_test = testset$data[, common_taxa, drop=F]
testset$meta = testset$meta %>% 
  mutate(disease = if_else(disease == "NGT", "Healthy", disease))
testset$meta$study = "testing_set"

```

### Generalization
```{r}
# re-train
ab.mint.splsda = tune(method = "mint.splsda", X = new_train, Y = as.factor(result_df$meta$disease), ncomp = 2, folds=5, nrepeat=10,
                      study = as.factor(result_df$meta$study), test.keepX = seq(10, ncol(new_train), 1), validation = 'Mfold', dist = "centroids.dist") 
plot(ab.mint.splsda)
optimal.keepX = ab.mint.splsda$choice.keepX

mint.splsda.model = mint.splsda(X=new_train, Y=result_df$meta$disease, 
                                study=result_df$meta$study, ncomp=2, keepX = optimal.keepX)

perf(mint.splsda.model, validation="Mfolds", folds=5, nrepeat=10, dist="centroids.dist")$global.error$error.rate.class


## Confusion matrix for testing set
predict.splsda.null <- predict(mint.splsda.model, newdata = new_test, 
                             dist = "centroids.dist", study.test = testset$meta$study)

conf_mat_c1 = get.confusion_matrix(truth = testset$meta$disease, predicted = predict.splsda.null$class$centroids.dist[, 1])
conf_mat_c2 = get.confusion_matrix(truth = testset$meta$disease, predicted = predict.splsda.null$class$centroids.dist[, 2])
conf_mat_c1
conf_mat_c2
```

## train model on overlap data frame and use testing set to validate performance
## make sure they have the same columns 
training=a, b
testing a/b --> maybe include this as two columns?
## look at confusion matrix





