

```{r}
# import libraries
library(testthat)
library(dplyr)
library(SummarizedExperiment)
library(mixOmics)
library(stringr)
library(tibble)
library(stringr)
library(purrr)
library(curatedMetagenomicData)
```
## Filtering and Normalizing Metadata

The function `filtered_meta_list()` takes a metadata dataframe and applies 
a mapping of regex patterns to normalize disease labels.  

**Arguments:**
- `meta`: Dataframe with a column `"disease"`. Row names are sample IDs.  
- `disease_map`: Named list where regex patterns map to standardized labels.  
- `cols_to_keep`: List of columns from metadata to be kept

**Returns:**  
A tibble with:  
- `SID`: Sample IDs (from rownames)  
- `disease`: Normalized disease states 
- `metadata_colums`: Other columns from metadata to be kept
```{r}

filter_meta <- function(meta, diseases, cols_to_keep = NULL) {
  
  # 1. Check input
  if (!"disease" %in% colnames(meta)) {
    stop("`meta` must contain a column named 'disease'.")
  }
  if (!is.null(cols_to_keep)) {
    missing_cols <- setdiff(cols_to_keep, colnames(meta))
    if (length(missing_cols) > 0) {
      stop("The following columns are missing from metadata: ",
           paste(missing_cols, collapse = ", "))
    }
  }
  
  # 2. Normalize disease names (case-insensitive)
  meta <- meta %>%
    mutate(disease = map_chr(disease, function(x) {
      matched <- NA
      for (pattern in diseases) {
        if (str_detect(x, regex(pattern, ignore_case = TRUE))) {
          matched <- pattern
          break
        }
      }
      if (is.na(matched)) x else matched
    }))
  
  # 3. Filter metadata
  filtered_meta <- meta %>% filter(disease %in% diseases)
  
  # 4. Select columns
  if (!is.null(cols_to_keep)) {
    filtered_meta <- filtered_meta %>% dplyr::select(disease, all_of(cols_to_keep))
  } else {
    filtered_meta <- filtered_meta %>% dplyr::select(disease)
  }
  
  # 5. Return (row names preserved automatically by R)
  return(filtered_meta)
}

```

## Split Count Filtering by Disease

The function `split_count_filtering()` filters an abundance dataframe by applying
low-abundance OTU removal **within each disease group** and then taking the union
of retained OTUs across all groups. If there is only one disease, it simply applies
low-count removal to the full dataset.

**Arguments:**
- `labelled_abund`: Abundance dataframe/tibble with `SID` as the first column.
- `diseases`: Vector of disease names to split by.
- `percent`: Numeric threshold for low count removal (same as in `low_count_removal`).
- `meta`: Metadata dataframe containing `SID` and `disease` columns (needed to subset samples by disease).

**Returns:**
- A filtered abundance dataframe containing only OTUs that survived low-count filtering
  across all specified disease groups.
```{r}
split_count_filtering <- function(labelled_abund, diseases, percent, filtered_meta) {
  
  # Case: less than 2 diseases
  if (length(diseases) < 2) {
    message("Not enough diseases to split; running low count removal on full dataset")
    res <- low_count_removal(labelled_abund, percent)
    abundance_filtered <- res$data.filter
  } else {
    all_keep_otu <- c()
    
    for (disease_name in diseases) {
      
      # Subset samples matching this disease, preserving rownames
      disease_sids <- rownames(filtered_meta)[filtered_meta$disease == disease_name]
      disease_df <- labelled_abund[disease_sids, , drop = FALSE]
    
      
      # Apply low count removal
      res <- low_count_removal(disease_df, percent)
      
      all_keep_otu <- union(all_keep_otu, res$keep.otu)
    }
    
    # Filter original abundance table by OTUs that survived across all groups
    
    abundance_filtered <- labelled_abund[, all_keep_otu, drop = FALSE]
    
    # Ensure numeric matrix, no lists, preserve rownames
    abundance_filtered <- as.data.frame(lapply(abundance_filtered, function(col) {
      if (is.list(col)) col <- unlist(col)
      as.numeric(col)
    }), row.names = rownames(abundance_filtered))
  }
  #View(abundance_filtered)
  return(abundance_filtered)
}
```

## Remove low-abundance OTUs

The function `low_count_removal()` filters OTUs (columns) in a count matrix or dataframe
based on their relative abundance. Only OTUs that contribute more than a specified
percentage of the total counts are kept.

**Arguments:**
- `data`: Dataframe or matrix of OTU counts (n samples × p OTUs)
- `percent`: Numeric threshold (default `0.0001`). OTUs contributing more than this
  percentage of total counts are kept.

**Returns:**
A list with two elements:
- `data.filter`: Filtered count dataframe containing only OTUs above the threshold
- `keep.otu`: Indices of the OTUs that were kept
```{r}
low_count_removal <- function(
  data,    # OTU count dataframe of size n (samples) x p (OTUs)
  percent = 0.0001  # cutoff percentage threshold
) {
  
  # Check input type
  if (!is.data.frame(data) && !is.matrix(data)) {
    stop("`data` must be a dataframe or matrix of counts.", call. = FALSE)
  }
  
  if (!all(sapply(data, is.numeric))) stop("All columns must be numeric.")

  
  # Identify OTUs above the relative abundance cutoff
  keep.otu <- which(
    colSums(data) * 100 / sum(colSums(data)) > percent
  )
  
  # Filter the dataframe to keep only those OTUs
  data.filter <- data[, keep.otu, drop = FALSE]
  
  # Return both filtered data and indices
  return(list(
    data.filter = data.filter,
    keep.otu = keep.otu
  ))
}
```

## CLR Transformation of Abundance Data

The function `clr.transformation()` applies a **centered log-ratio (CLR) transformation**
to an abundance dataframe using the `logratio.transfo()` function from the **MixOmics** package.

CLR is commonly used in compositional data analysis to transform relative abundance data
into a form suitable for downstream analyses.

**Arguments:**
- `abundance.filtered`: Dataframe or matrix of filtered abundance values (OTUs as columns, samples as rows)
- `offset_value`: Numeric value added to each entry to avoid taking log of zero

**Returns:**
- A dataframe of the CLR-transformed abundance values, with the same dimensions as input.
```{r}
clr.transformation <- function(abundance.filtered, offset_value) {
  # Apply CLR transformation using MixOmics
  # logratio.transfo requires a matrix input
  data.clr <- logratio.transfo(
    as.matrix(abundance.filtered),
    logratio = "CLR",
    offset = offset_value
  )
  
  # Convert result to a regular dataframe
  clr_mat <- unclass(data.clr)
  clr_df <- as.data.frame(clr_mat)
  
  return(clr_df)
}

```


## Preprocessing of Single Study Data

The function `preprocess_single()` performs **preprocessing of a single study’s microbiome dataset**

It filters, normalizes, and structures the input abundance and metadata tables to ensure consistency across datasets before multi-study integration.

**Arguments:**
- `meta`: Dataframe containing metadata. Rows correspond to samples, and columns include variables such as disease labels and other covariates.  
- `abund`: Dataframe of raw or relative abundance data (samples as rows, taxa/OTUs as columns).  
- `diseases`: Character vector specifying which disease groups to include for analysis. Samples with other disease labels are excluded.  
- `keep_cols`: Character vector of metadata column names to retain in the output metadata.  
- `percent`: Numeric value (0–100) specifying the threshold for **low-count removal** — taxa present in fewer than this percentage of samples are filtered out.  
- `offset_value`: Numeric constant added to abundance values prior to **log or CLR transformation** to prevent taking the log of zero.

**Returns:**
Returns two options:
If CLR = TRUE (default)
- A **list** containing two elements:
  - `$data`: The processed abundance dataframe after filtering, normalization, and transformation.  
  - `$meta`: The corresponding filtered metadata dataframe aligned with `$data`.  
If CLR = FALSE
- A **matrix** containing disease selected abundance that has been split filtered. 
```{r}
preprocess_single <- function(meta, abund, diseases, keep_cols, percent, offset_value, CLR=TRUE) {
  # 1. Filter metadata
  filtered_meta <- filter_meta(meta, diseases, keep_cols)
  #View(filtered_meta)
  # 2. Select abundance rows that match filtered_meta
  selected_abund <- as.data.frame(abund[rownames(abund) %in% rownames(filtered_meta), ])
  #View(selected_abund)
  # 3. Apply split count filtering
  abundance_filtered <- split_count_filtering(selected_abund, diseases, percent, filtered_meta)
  
 if (!CLR) return(abundance_filtered)
  # 5. Convert to matrix for CLR
  abundance_matrix <- as.matrix(abundance_filtered)
  
  # 6. Apply CLR transformation
  clr_transformed <- clr.transformation(abundance_matrix, offset_value)
  #View(clr_transformed)
  # 7. Return results
  clr_transformeddf <- as.data.frame(clr_transformed)
      return(list(
    data = clr_transformeddf,
    meta = filtered_meta
  ))

}
```
## Preprocessing Multiple Studies

The function `preprocess_multiple()` applies `preprocess_single()` to each study in a list
and combines the results into a single dataframe.

**Arguments:**
- `abundance_list`: Named list of abundance dataframes (samples × taxa).  
- `meta_list`: Named list of metadata dataframes.  
- `diseases`: Character vector of disease groups to include.  
- `keep_cols`: Metadata columns to retain.  
- `percent`: Threshold for low-count taxa removal.  
- `offset_value`: Numeric offset for CLR/log transformation.  
- `threshold`: Minimum number of studies a taxon must appear in to be retained.  

**Returns:**
- Combined dataframe with all samples, filtered taxa, metadata, and study.

```{r}
preprocess_multiple <- function(abundance_list, meta_list, diseases, keep_cols, percent, offset_value, threshold = NULL) {
  results <- list()
  # iterate through abund and meta lists to proprocess individually
  for (nm in names(abundance_list)) {
    abund <- abundance_list[[nm]]
    meta  <- meta_list[[nm]]
    result <- preprocess_single(meta, abund, diseases, keep_cols, percent, offset_value)
    
    # store it inside results under the dataset named results accessed by name
    results[[nm]] <- list(
      data = result$data,
      meta = result$meta
    )
    cat("\n--- Dataset:", nm, "---\n")
  print(dim(result$data))
  # View(result$data)
  # View(result$meta)
  }
  
  combined <- integrate_filter(results, threshold)
  return(combined)
}

```

## Integrate Filtered Abundance Across Studies

The function `integrate_filter()` combines multiple preprocessed studies into a single dataframe,
retaining only taxa that appear in at least `threshold` studies.

**Arguments:**
- `results`: Named list of preprocessed study results, each with:
  - `$data`: Abundance dataframe (samples × taxa)  
  - `$meta`: Metadata dataframe  
- `threshold`: Minimum number of studies a taxon must appear in to be retained.  

**Returns:**
- A combined dataframe containing filtered abundance, metadata, and a `study` column indicating the source study.

```{r}
integrate_filter <- function(results, threshold) {
  
  # Step 1: Determine taxa present above threshold across studies
  taxa_present_list <- purrr::map(results, function(study) {
    taxa_cols <- colnames(study$data)
    present_taxa <- taxa_cols[
      purrr::map_lgl(taxa_cols, ~ sum(study$data[[.x]], na.rm = TRUE) > 0)
    ]
    return(present_taxa)
  })
  
  taxa_counts <- table(unlist(taxa_present_list))
  surviving_taxa <- names(taxa_counts)[taxa_counts >= threshold]
  
  # Step 2: Filter each study
  comb_data_list <- list()
  meta_list <- list()
  
  
  
  
final <- purrr::imap(results, function(study, study_name) {
  
  # Filter abundance to surviving taxa
  abund_filtered <- study$data %>%
    dplyr::select(any_of(surviving_taxa))
  
  # Preserve rownames
  rownames(abund_filtered) <- rownames(study$data)
  
  # Filter metadata to match abundance rows
  meta_filtered <- study$meta[rownames(abund_filtered), , drop = FALSE]
  
  # Return a nested list, automatically named by imap using study_name
  list(
    abund = abund_filtered,
    meta = meta_filtered
  )
})

 
  return(final)
}
```

## Run Preprocessing on Single or Multiple Datasets

The function `run_preprocessing()` automatically processes one or more datasets.  
If a single dataset is provided, it calls `preprocess_single()`.  
If multiple datasets are provided, it calls `preprocess_multiple()` and integrates them.

**Arguments:**
- `abundance_list`: Named list of abundance dataframes (samples × taxa)  
- `meta_list`: Named list of corresponding metadata dataframes  
- `diseases`: Character vector of disease labels to retain  
- `keep_cols`: Metadata columns to keep  
- `percent`: Low-count filtering threshold (percent)  
- `offset_value`: Numeric offset for CLR transformation  
- `threshold`: (Optional) Minimum number of studies a taxon must appear in (for multiple datasets)  

**Returns:**
- A processed abundance dataframe (CLR-transformed) combined with metadata.  
```{r}
run_preprocessing <- function(abundance_list, meta_list, diseases,
                              keep_cols, percent, offset_value, threshold = NULL) {
  # Check how many datasets are in the abundance list
  if (length(abundance_list) == 1) {
    print("single")
    # Only one dataset → run preprocess_single
    #how to access this type of database
    abund <-  as.data.frame((abundance_list)[[1]])
    meta  <- as.data.frame((meta_list)[[1]])
    result <- preprocess_single(meta, abund, diseases, keep_cols, percent, offset_value)
    
  } else {
    print("Multiple")
    # Multiple datasets → run preprocess_combine (preprocess_multiple)
    result <- preprocess_multiple(abundance_list, meta_list,
                                 diseases, keep_cols, percent, offset_value, threshold)
  }
  return(result)
}
```





## Testing with single dataset
```{r}
#testing with actual data
rawdata <- curatedMetagenomicData("SankaranarayananK_2015.relative_abundance", dryrun = FALSE, counts = TRUE,rownames = "short"  )
```

```{r}
## from treesummarisedexperiment data structure obtain raw_tse and abudance
raw_tse <- rawdata[[1]]
abund1 <- assay(raw_tse, "relative_abundance")

# from raw_tse obtain meta data as dataframe
meta1 <- as.data.frame(colData(raw_tse))

# transcribe the dataframe abund so now the column names are row names and vice versa
abund1_t <- as.data.frame(t(abund1))
diseases <- c("Healthy", "T2D")
keep_cols <- c("gender")
result<- preprocess_single(meta = meta1, abund = abund1_t, diseases = diseases, keep_cols = keep_cols, percent = 0.0001, offset_value = 1, CLR=TRUE)
```





## testing with 6 datasets from paper
```{r}
# List of study names
study_names <- c("FengQ_2015.relative_abundance", "HMP_2019_t2d.relative_abundance", 
                 "KarlssonFH_2013.relative_abundance", "LiJ_2014.relative_abundance", 
                 "YuJ_2015.relative_abundance", "SankaranarayananK_2015.relative_abundance")

# Load and process each study
abundance_lst <- lapply(study_names, function(study) {
  rawdata <- curatedMetagenomicData(study, dryrun = FALSE, counts = TRUE, rownames = "short")
  raw_tse <- rawdata[[1]]
  abund <- assay(raw_tse, "relative_abundance")
  as.data.frame(t(abund))
})
names(abundance_lst) <- study_names

# Extract metadata for each study
meta_lst <- lapply(study_names, function(study) {
  rawdata <- curatedMetagenomicData(study, dryrun = FALSE, counts = TRUE, rownames = "short")
  raw_tse <- rawdata[[1]]
  as.data.frame(colData(raw_tse))
})
names(meta_lst) <- study_names
keep_cols <-c("gender", "body_site")
diseases <- c("Healthy", "T2D")
```


```{r}
res<- run_preprocessing(abundance_list = abundance_lst, meta_list = meta_lst, diseases = diseases, keep_cols = keep_cols, percent = 0.01, offset_value = 1, threshold = 1)

saveRDS(res, "res")
```
```{r}
#exporting code
saveRDS(res, "df")
```




## saving datasets for testing the original modelling code
```{r}
#testing with actual data
rawdata1 <- curatedMetagenomicData("SankaranarayananK_2015.relative_abundance", dryrun = FALSE, counts = TRUE,rownames = "short"  )
```

```{r}
## from treesummarisedexperiment data structure obtain raw_tse and abudance
raw_tse1 <- rawdata1[[1]]


abund1 <- assay(raw_tse1, "relative_abundance")

# from raw_tse obtain meta data as dataframe
meta1 <- as.data.frame(colData(raw_tse1))

# transcribe the dataframe abund so now the column names are row names and vice versa
abund1_t <- as.data.frame(t(abund1))
diseases <- c("Healthy", "T2D")
keep_cols <- c("gender")
result1<- preprocess_single(meta = meta1, abund = abund1_t, diseases = diseases, keep_cols = keep_cols, percent = 0.0001, offset_value = 1, CLR=TRUE)


saveRDS(result1, "sank_clr")
```

```{r}
#testing with actual data
rawdata2 <- curatedMetagenomicData("FengQ_2015.relative_abundance", dryrun = FALSE, counts = TRUE,rownames = "short"  )
```

```{r}
## from treesummarisedexperiment data structure obtain raw_tse and abudance
raw_tse2 <- rawdata2[[1]]
abund2 <- assay(raw_tse2, "relative_abundance")

# from raw_tse obtain meta data as dataframe
meta2 <- as.data.frame(colData(raw_tse2))

# transcribe the dataframe abund so now the column names are row names and vice versa
abund2_t <- as.data.frame(t(abund2))
diseases <- c("Healthy", "T2D")
keep_cols <- c("gender")
result2<- preprocess_single(meta = meta2, abund = abund2_t, diseases = diseases, keep_cols = keep_cols, percent = 0.0001, offset_value = 1, CLR=TRUE)
saveRDS(result2$data, "feng_clr")
```