---
title: "Individual_filtering"
output: html_document
---


```{r}
library(mixOmics)
library(dplyr)
```

## Filter studies by PLS-DA misclassification

The function `individual_filtering()` evaluates one or more microbiome studies based on
their cross-validated PLS-DA performance. Studies (or single datasets) are retained only
if their class-level misclassification rates do not exceed a specified threshold.

**Arguments:**
- `data`: A single study (list with `abund` and `meta`) or a nested list of studies.  
  Each study must contain:  
  - `abund`: Dataframe or matrix of abundances (n samples × p features)  
  - `meta`: Dataframe with sample metadata, including a `disease` column for outcomes
- `err_thres`: Numeric threshold for class misclassification (default `0.7`).  
  Studies with any sample exceeding this error are filtered out.
- `ncomp`: Number of components for the PLS-DA model (default `2`)
- `validation`: Type of cross-validation (default `"Mfold"`)
- `folds`: Number of folds for cross-validation (default `5`)
- `dist`: Distance metric used in PLS-DA performance (default `"max.dist"`)

**Returns:**
- If a single study is provided: `TRUE` if it passes the threshold, `FALSE` otherwise
- If a nested list of studies is provided: A filtered list containing only studies that
  passed the threshold, keeping the original structure (`abund` + `meta`)
```{r}
individual_filtering <- function(data, err_thres = 0.7, ncomp = 2,
                                 validation = "Mfold", folds = 5, dist = "max.dist") {
  
  # Helper to check a single study
  check_single_dataset <- function(study) {
    abund <- study$abund   # renamed from bund
    meta  <- study$meta
    Y <- factor(meta$disease)
    
    plsda.model <- plsda(abund, Y, ncomp = ncomp)
    perf.model <- perf(
      object = plsda.model,
      validation = validation,
      folds = folds,
      progressBar = FALSE,
      dist = dist
    )
    
    class_errors <- perf.model$error.rate.class[[dist]]
    exceed_threshold <- apply(class_errors, 1, function(x) any(x > err_thres))
    
    keep_dataset <- !any(exceed_threshold)
    return(keep_dataset)
  }
  
  # Case 1: single study (has both abund and meta)
  if (is.list(data) && all(c("abund","meta") %in% names(data))) {
    return(check_single_dataset(data))
  }
  
  # Case 2: nested list of studies
  if (is.list(data) && !all(c("abund","meta") %in% names(data))) {
    results <- purrr::imap(data, function(study, study_name) {
      if (!all(c("abund","meta") %in% names(study))) {
        stop(paste("Study", study_name, "is missing abund or meta"))
      }
      keep <- check_single_dataset(study)
      if (keep) return(study) else return(NULL)
    })
    results <- results[!purrr::map_lgl(results, is.null)]
    return(results)
  }
  
  stop("Input must be a single study (abund + meta) or a nested list of studies")
}



```


```{r}
# individual filtering testing
CMD_filtered <- readRDS("res")

# test with a list of studies --> returned a list with all studies except single_lij
result <- individual_filtering(CMD_filtered)

# test with a single study --> returned FALSE
single_lij <- individual_filtering(response$LiJ_2014.relative_abundance)

```




## My test
```{r}
check_single_dataset <- function(study) {
    abund <- study$abund   # renamed from bund
    meta  <- study$meta
    Y <- factor(meta$disease)
    
    plsda.model <- plsda(abund, Y, ncomp = ncomp)
    perf.model <- perf(
      object = plsda.model,
      validation = validation,
      folds = folds,
      progressBar = FALSE,
      dist = dist
    )
    
    #class_errors <- perf.model$error.rate.class[[dist]]
    #exceed_threshold <- apply(class_errors, 1, function(x) any(x > err_thres))
    class_errors <- perf.model$error.rate.class[[dist]][, ncomp]
    exceed_threshold <- any(class_errors > err_thres)
    
    keep_dataset <- !any(exceed_threshold)
    return(keep_dataset)
}
CMD_filtered <- readRDS("res")

```


```{r}
i = 6
names(CMD_filtered)[i]
abund=CMD_filtered[[i]]$abund
meta=CMD_filtered[[i]]$meta
Y=factor(meta$disease)
ncomp=2 ; validation="Mfold" ; folds=5 
dist = "max.dist"
dist="centroids.dist"
err_thres=0.7

plsda.model <- plsda(abund, Y, ncomp = ncomp)
    perf.model <- perf(
      object = plsda.model,
      validation = validation,
      folds = folds,
      progressBar = FALSE,
      dist = dist
    )

class_errors <- perf.model$error.rate.class[[dist]][, ncomp]
# exceed_threshold <- apply(class_errors, 2, function(x) any(x > err_thres))
exceed_threshold <- any(class_errors > err_thres)
!any(exceed_threshold)
```


```{r}

```



